---
title: "AI by hand - lectures by Dr. Tom Yeh"
subtitle: "Vector databases"
author: "Posit Assistant, guided by Dan Killian"
date: 2-27-2026
toc: true
toc-depth: 3
number-sections: true
format:
  html:
    code-fold: true
    page-layout: full
execute: 
  keep-md: true
  warning: false
  message: false
editor: visual
reference-location: margin
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../prep (May 2025).r")
# sna (loaded by prep) masks gt::gt() — pin the correct one to the global env
library(ggrepel)
gt <- gt::gt
```

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

## Introduction

This tutorial walks through how a **vector database** works from first principles, following the structure of Dr. Tom Yeh's [Vector Database by Hand](https://www.byhand.ai/p/14-can-you-calculate-a-vector-database). Every computation is performed explicitly in R so the mechanics are fully visible.

Vector databases are the backbone of **Retrieval Augmented Generation (RAG)** — the technique that allows large language models to search and retrieve relevant information from a corpus before generating a response. Instead of storing data as rows and columns, a vector database stores data as numeric vectors called *embeddings*, and retrieves information by geometric similarity rather than exact match.

The pipeline has four stages:

| Stage | Input | Output |
|---|---|---|
| Word Embeddings | Words (text) | Numeric vectors per word |
| Encoding + Pooling | Word vectors | Single vector per sentence |
| Indexing | Full-dim vectors | Reduced-dim vectors stored in DB |
| Retrieval | Query vector | Most similar item from DB |

We will work through each stage with a tiny corpus of three sentences:

- **"how are you"**
- **"who are you"**
- **"who am I"**

And we will query the database with: **"am I you"**

---

## Data

Our corpus has 3 sentences, each 3 words long. We define a vocabulary of the 6 unique words that appear across all sentences and the query.

```{r corpus}
sentences <- list(
  s1 = c("how", "are", "you"),
  s2 = c("who", "are", "you"),
  s3 = c("who", "am",  "I")
)

query_words <- c("am", "I", "you")

vocab <- c("how", "are", "you", "who", "am", "I")

tibble(
  Sentence = c("how are you", "who are you", "who am I"),
  Label    = c("s1", "s2", "s3")
) |>
  gt() |>
  tab_header(title = "Corpus") |>
  cols_label(Sentence = "Text", Label = "ID")
```

In practice, a corpus may contain millions of documents with a vocabulary of tens of thousands of tokens. The math is the same — just bigger.

---

## Word Embeddings

The first step converts each word into a numeric vector. These vectors are stored in a **word embedding table**: a matrix where each column is a word and each row is one dimension of the embedding space.

The values in this table are learned during model training to capture semantic relationships. Here we use a hand-crafted 4-dimensional table (values drawn from Dr. Yeh's spreadsheet where available) for illustration:

```{r word-embeddings}
E <- matrix(
  c(# how  are  you  who   am    I
     1,   0,   0,   1,   0,   0,   # dim 1  (wh-question words)
     0,   1,   0,   0,   1,   0,   # dim 2  (verb-like words)
     1,   1,   0,   0,   0,   1,   # dim 3  (sentence-structure words)
     0,   0,   1,   0,   1,   1    # dim 4  (person/object words)
  ),
  nrow = 4, byrow = TRUE,
  dimnames = list(paste0("dim", 1:4), vocab)
)

E |>
  as.data.frame() |>
  rownames_to_column("Dimension") |>
  gt() |>
  tab_header(
    title    = "Word Embedding Table E",
    subtitle = "4 dimensions × 6 vocabulary words"
  ) |>
  tab_spanner(label = "Vocabulary", columns = all_of(vocab)) |>
  data_color(
    columns = all_of(vocab),
    method  = "numeric",
    palette = c("white", "#A7C6ED", "#002F6C")
  )
```

To embed a sentence, we simply **look up** each word's column from this table. For "how are you":

```{r lookup-example}
sentence1_embs <- E[, sentences$s1]

sentence1_embs |>
  as.data.frame() |>
  rownames_to_column("Dimension") |>
  gt() |>
  tab_header(
    title    = "Word Lookup: \"how are you\"",
    subtitle = "Each column is retrieved from the embedding table E"
  ) |>
  cols_label(how = "how", are = "are", you = "you")
```

> **Key insight:** At this point we have a *matrix* of embeddings — one column per word. We need a single vector to represent the whole sentence. That is the job of the encoder.

---

## Text Embeddings: Encoding and Pooling

Turning a variable-length sequence of word vectors into a single fixed-size vector requires two sub-steps: **encoding** and **pooling**.

### Linear Encoder + ReLU {.unnumbered}

The encoder applies a weight matrix **W** to each word embedding, followed by a ReLU activation (which sets any negative values to zero). This is a single-layer perceptron — the simplest possible neural network encoder. In practice, a full transformer architecture is used.

```{r encoder}
# Weight matrix W (4×4): learned during training, hand-crafted here
W <- matrix(
  c( 1,  0,  1,  0,
     0,  1,  0,  1,
     1,  0,  0,  1,
     0,  1,  1,  0),
  nrow = 4, byrow = TRUE,
  dimnames = list(paste0("f", 1:4), paste0("dim", 1:4))
)

W |>
  as.data.frame() |>
  rownames_to_column("Feature") |>
  gt() |>
  tab_header(
    title    = "Encoder Weight Matrix W",
    subtitle = "4×4: maps word embeddings to feature vectors"
  ) |>
  data_color(
    columns = starts_with("dim"),
    method  = "numeric",
    palette = c("white", "#A7C6ED", "#002F6C")
  )
```

The ReLU function is applied element-wise after the matrix multiplication:

$$\text{ReLU}(x) = \max(0, x)$$

```{r relu-demo}
relu <- function(x) pmax(x, 0)

# Show encoding of "how are you" word by word
linear_out <- W %*% sentence1_embs
relu_out   <- relu(linear_out)

list(
  `Linear (W × embeddings)` = linear_out,
  `After ReLU`              = relu_out
) |>
  lapply(\(m) as.data.frame(m) |> rownames_to_column("Feature")) |>
  bind_rows(.id = "Step") |>
  gt(groupname_col = "Step") |>
  tab_header(
    title    = "Encoding \"how are you\"",
    subtitle = "Step 1: W × word_embedding | Step 2: ReLU(result)"
  ) |>
  fmt_number(decimals = 2) |>
  data_color(
    columns = where(is.numeric),
    method  = "numeric",
    palette = c("white", "#A7C6ED", "#002F6C")
  )
```

### Mean Pooling {.unnumbered}

We now have one feature vector per word. Mean pooling collapses them into a **single sentence vector** by averaging across the word dimension:

$$\text{pooled} = \frac{1}{n} \sum_{i=1}^{n} \text{feature}_i$$

```{r mean-pooling}
pooled_s1 <- rowMeans(relu_out)

tibble(
  Feature = paste0("f", 1:4),
  `ReLU(how)` = relu_out[, "how"],
  `ReLU(are)` = relu_out[, "are"],
  `ReLU(you)` = relu_out[, "you"],
  `Mean (pooled)` = pooled_s1
) |>
  gt() |>
  tab_header(
    title    = "Mean Pooling: \"how are you\"",
    subtitle = "Average across words to produce one sentence vector"
  ) |>
  fmt_number(decimals = 3) |>
  tab_style(
    style = cell_fill(color = "#A7C6ED"),
    locations = cells_body(columns = `Mean (pooled)`)
  )
```

The pooled vector is the **text embedding** (also called a sentence embedding). It represents the whole sentence in 4-dimensional space.

---

## Indexing

Storing and comparing full-dimensional vectors at scale is expensive. **Indexing** reduces the embedding to a shorter vector using a **projection matrix P**, similar in spirit to hashing — trading some precision for speed.

Here we project from 4 dimensions down to 2:

```{r projection}
P <- matrix(
  c(1, 0, 0, 1,
    0, 1, 1, 0),
  nrow = 2, byrow = TRUE,
  dimnames = list(paste0("p", 1:2), paste0("f", 1:4))
)

P |>
  as.data.frame() |>
  rownames_to_column("Index dim") |>
  gt() |>
  tab_header(
    title    = "Projection Matrix P",
    subtitle = "2×4: compresses 4-d pooled vector to 2-d index"
  ) |>
  data_color(
    columns = starts_with("f"),
    method  = "numeric",
    palette = c("white", "#A7C6ED", "#002F6C")
  )
```

Now we run the complete pipeline for all three sentences and store their 2-d index vectors:

```{r build-database}
# Full encoding pipeline
encode_sentence <- function(words) {
  word_embs <- E[, words, drop = FALSE]     # word embedding lookup
  features  <- relu(W %*% word_embs)        # linear encoder + ReLU
  pooled    <- rowMeans(features)            # mean pooling
  index     <- drop(P %*% pooled)           # projection to 2-d
  list(word_embs = word_embs,
       features  = features,
       pooled    = pooled,
       index     = index)
}

corpus_results <- lapply(sentences, encode_sentence)

# Assemble the vector database (index vectors only)
db <- do.call(rbind, lapply(corpus_results, `[[`, "index"))
rownames(db) <- c("how are you", "who are you", "who am I")
colnames(db) <- c("p1", "p2")

db |>
  as.data.frame() |>
  rownames_to_column("Sentence") |>
  gt() |>
  tab_header(
    title    = "Vector Database",
    subtitle = "2-dimensional index vectors for each sentence"
  ) |>
  fmt_number(decimals = 3) |>
  data_color(
    columns = c(p1, p2),
    method  = "numeric",
    palette = c("white", "#A7C6ED", "#002F6C")
  )
```

We can visualize the three sentences as points in 2-d index space:

```{r plot-db, fig.height=4, fig.width=5}
db |>
  as.data.frame() |>
  rownames_to_column("Sentence") |>
  ggplot(aes(x = p1, y = p2, label = Sentence)) +
  geom_point(size = 4, color = "#002F6C") +
  ggrepel::geom_label_repel(size = 3.5, fill = "#A7C6ED",
                             box.padding = 0.5, point.padding = 0.3) +
  labs(
    title    = "Vector Database: Index Space",
    subtitle = "Each sentence occupies a unique position in 2-d space",
    x = "Index dimension 1 (p1)",
    y = "Index dimension 2 (p2)"
  )
```

---

## Retrieval

To query the database, we process the query phrase **"am I you"** through the identical pipeline to obtain a 2-d query vector. We then compute the **dot product** between the query vector and every stored index vector. The sentence with the highest dot product is returned as the nearest neighbor.

### Encoding the query {.unnumbered}

```{r encode-query}
q_result <- encode_sentence(query_words)

tibble(
  Step   = c("Query words", "Index vector (p1)", "Index vector (p2)"),
  Value  = c(
    paste(query_words, collapse = " → "),
    round(q_result$index[1], 3),
    round(q_result$index[2], 3)
  )
) |>
  gt() |>
  tab_header(title = "Query Pipeline: \"am I you\"")
```

### Dot product similarity {.unnumbered}

The dot product measures how aligned two vectors are. Higher value = more similar.

$$\text{similarity}(q, d) = q \cdot d = \sum_{i} q_i \cdot d_i$$

```{r dot-products}
q_vec <- q_result$index
dots  <- drop(db %*% q_vec)

tibble(
  Sentence       = names(dots),
  `p1`           = db[, "p1"],
  `p2`           = db[, "p2"],
  `Dot product`  = dots,
  `Nearest neighbor` = ifelse(dots == max(dots), "✓", "")
) |>
  arrange(desc(`Dot product`)) |>
  gt() |>
  tab_header(
    title    = "Retrieval: Dot Product vs. Query \"am I you\"",
    subtitle = paste0("Query index vector: [", round(q_vec[1], 3),
                      ", ", round(q_vec[2], 3), "]")
  ) |>
  fmt_number(columns = c(p1, p2, `Dot product`), decimals = 3) |>
  tab_style(
    style     = list(cell_fill(color = "#A7C6ED"), cell_text(weight = "bold")),
    locations = cells_body(rows = `Nearest neighbor` == "✓")
  )
```

### Visualization {.unnumbered}

```{r plot-retrieval, fig.height=4.5, fig.width=5.5}
db_df <- db |>
  as.data.frame() |>
  rownames_to_column("Sentence") |>
  mutate(type = "Database")

query_df <- data.frame(
  Sentence = "am I you (query)",
  p1 = q_vec[1], p2 = q_vec[2],
  type = "Query"
)

bind_rows(db_df, query_df) |>
  ggplot(aes(x = p1, y = p2, color = type, label = Sentence)) +
  geom_point(size = 4) +
  ggrepel::geom_label_repel(size = 3.2, box.padding = 0.6, point.padding = 0.3,
                             fill = "white") +
  scale_color_manual(values = c(Database = "#002F6C", Query = "#BA0C2F")) +
  labs(
    title    = "Query vs. Database in Index Space",
    subtitle = "\"am I you\" (red) is closest to \"who am I\" (blue)",
    x = "Index dimension 1 (p1)",
    y = "Index dimension 2 (p2)",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

**Result:** The nearest neighbor to "am I you" is **"who am I"** — geometrically, these sentences share the most similar structure (same person-referent words: *am*, *I*, *you*).

---

## From "By Hand" to Practice

The toy example above captures the complete logic of a production vector database. The main differences at scale are:

| This tutorial | Production system |
|---|---|
| 6-word vocabulary, 4-d embeddings | 50k+ token vocabulary, 1536-d embeddings |
| Hand-crafted weight matrix W | Transformer encoder (e.g., BERT, text-embedding-3-small) |
| 3-sentence corpus | Millions of documents |
| Exact linear scan (argmax) | Approximate Nearest Neighbor (HNSW, IVF) |
| 2-d index via hand-crafted P | Learned index (e.g., FAISS, pgvector) |

### Using real embeddings in R {.unnumbered}

The `ragnar` package provides a direct bridge to OpenAI's embedding API, which produces 1536-dimensional embeddings using a transformer encoder. The same retrieval logic (cosine similarity = normalized dot product) applies.

```{r ragnar-demo, eval=FALSE}
library(ragnar)

# Generate embeddings via OpenAI text-embedding-3-small
emb_mat <- ragnar::embed_openai(
  text       = c("how are you", "who are you", "who am I"),
  model      = "text-embedding-3-small",
  batch_size = 10L
)

# L2-normalize so cosine similarity = dot product
row_l2 <- function(M) sqrt(rowSums(M * M))
Xn     <- emb_mat / row_l2(emb_mat)

# Query
q_raw <- ragnar::embed_openai("am I you", model = "text-embedding-3-small")
qn    <- drop(q_raw) / sqrt(sum(q_raw^2))

# Nearest neighbor
sims <- drop(Xn %*% qn)
sentences[which.max(sims)]
```

The vector database community has also produced dedicated stores — [pgvector](https://github.com/pgvector/pgvector) (PostgreSQL extension), [Pinecone](https://www.pinecone.io/), [Chroma](https://www.trychroma.com/), and [Weaviate](https://weaviate.io/) — that handle indexing, persistence, and approximate nearest neighbor search at production scale.

---

## Summary

A vector database works in four steps:

1. **Word Embeddings** — look up a numeric vector for each word from a pre-trained table.
2. **Encoding + Pooling** — pass word vectors through a neural encoder (linear + ReLU), then average them into a single sentence vector.
3. **Indexing** — project to a lower-dimensional space and store; this makes retrieval fast.
4. **Retrieval** — encode the query the same way, compute dot products against all stored vectors, return the highest-scoring match.

The geometry does the semantic work: sentences that share words (or meaning, with a real encoder) end up close together in vector space, so the right answer rises to the top.
