---
title: "Vector database demo"
subtitle: "A riff on Dr. Yeh's tutorial [Vector Database by Hand](https://www.byhand.ai/p/14-can-you-calculate-a-vector-database)"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
source("../../prep (May 2025).R")
library(readr)
library(gt)
library(htmltools)
```

## Introduction

This is a tutorial on how vector databases work, inspired by Dr. Yeh's [Vector Database by Hand](https://www.byhand.ai/p/14-can-you-calculate-a-vector-database). We will explicitly follow the structure of that tutorial, but with our own data and possibly some deviations based on my own interests in what can be done with this tool. 

We will proceed with establishing a corpus of text data. We will then represent that text as a set of numeric vectors. These vectors are referred to as embeddings. Next, we will store those embeddings in a vector database. Finally, we will visualize and query the vector database to demonstrate its utility.

## Data

We will use a corpus of haikus as our core data. First, we will import the poems in a way we can look at them. 


```{r}
h <- read_csv("../../Poetry/haikus.csv") %>%
    select(text) %>%
    rownames_to_column("id") %>%
    mutate(haiku = str_replace_all(text, " / ", "<br>")) 

nrow((h))

```
We have nearly 50,000 haikus. Let's look at a few of them.

``` {r}

h_gt <- h %>%
    select(1,3) %>%
    filter(id %in% 1:4) %>%
    gt() %>%
    cols_width(id ~ px(20)) %>%
    fmt_markdown(haiku)

h_gt
```


For the embeddings, it could be a useful source of information to know both the poem and the line of the poem.  


```{r}
h_long <- h %>%
    separate_rows(text, sep=" / ") %>%
    group_by(id) %>%
    mutate(line = row_number()) %>%
    ungroup() %>%
    mutate(id2 = paste(id, line, sep="-")) %>%
    select(id, line, id2, haiku=text)

head(h_long)

```

We are not able to pass a vector of lines and a vector of poems into a common embedding space, because the two vectors have a different size. What we will do is compute line embeddings, and perhaps show at some point that we can recover poem embeddings by averaging the poem's line embeddings. 

```{r}

h_long_samp <- h_long %>%
    .[1:3000,]

library(ragnar)

# embeds returned as a numeric matrix: rows = lines, cols = embedding dims
emb_mat <- ragnar::embed_openai(
  h_long_samp$haiku,
  model = "text-embedding-3-small",
  batch_size = 10L
)

# attach as a list-column (one numeric vector per row)
h_long_emb <- h_long_samp |>
  mutate(vec = lapply(seq_len(nrow(emb_mat)), \(i) unname(emb_mat[i, ])))

```


```{r}

```


## Embeddings

## Vector database

## Visualization

## Query 


